# **Exploratory Data Analysis on what factors that can affect students' academic performance** {#title}

Copyright by Hafidz Firmansyah G. ([hafidzfg\@gmail.com](mailto:hafidzfg@gmail.com){.email}) as a submission for Tetris FastTrack Data Analytics Program by DQLab

-   Last Updated: 24-Mar-22

-   News article reference: <https://www.brookings.edu/blog/brown-center-chalkboard/2022/03/03/the-pandemic-has-had-devastating-impacts-on-learning-what-will-it-take-to-help-students-catch-up/>

-   Data source: <https://www.kaggle.com/datasets/impapan/student-performance-data-set>

    > P. Cortez and A. Silva. Using Data Mining to Predict Secondary School Student Performance. In A. Brito and J. Teixeira Eds., Proceedings of 5th FUture BUsiness TEChnology Conference (FUBUTEC 2008) pp. 5-12, Porto, Portugal, April, 2008, EUROSIS, ISBN 978-9077381-39-7.

# Introduction

## **Background**

> Found an article that mentions the impact of COVID19 to students' academic performance. At the end, the article suggests (citing government plans) that tutoring, expanded learning time, and summer learning programs can help students catch up in academic. This gave me an idea to explore the what factors that can affect students' grades.

## Hypothesis

1.  Study time and extra paid classes/tutor can affect students' academic performance

2.  Parents' education, occupation, cohabitation, and relationship quality with students can affect students' academic performance.

## Goals

> 1.  Find the correlation between parents' education, occupation, cohabitation status, and relationship quality with students to students' academic performance (grade).
> 2.  Find the correlation between study time, extra paid classes, and family educational support to students' academic performance

# Data Preparation

## **Load Library, Data, and Data Cleaning**

Make sure to install all the libraries if you haven't yet

```{r}
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("plotly")
install.packages("heplots")
install.packages("ggpubr")
install.packages("grid")
install.packages("gridExtra")
```

Load the libraries and early data cleaning

```{r}
#Load the library
library(tidyverse)
library(ggplot2)
library(plotly)
library(heplots)
library(ggpubr)
library(grid)
library(gridExtra)

#load the dataset
## I will only focus on the provided math subject dataset
student.data <- read.csv('./dataset/student/student-mat.csv', sep = ';')

#Make sure data format is correct
student.data$Medu <- as.character(student.data$Medu)
student.data$Fedu <- as.character(student.data$Fedu)
student.data$traveltime <- as.character(student.data$traveltime)
student.data$studytime <- as.character(student.data$studytime)
student.data$famrel <- as.character(student.data$famrel)
```

## **Take a peek at the data**

```{r}
colnames(student.data) #check the columns
head(student.data) 

#Check for NA data
apply(student.data, 2, function(x) any(is.na(x))) #no NA data found
```

### **Column information (taken from kaggle):**

    # Attributes for both student-mat.csv (Math course) and student-por.csv (Portuguese language course) datasets:
    ## No need to run anything here
    1 school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)
    2 sex - student's sex (binary: 'F' - female or 'M' - male)
    3 age - student's age (numeric: from 15 to 22)
    4 address - student's home address type (binary: 'U' - urban or 'R' - rural)
    5 famsize - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3)
    6 Pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart)
    7 Medu - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
    8 Fedu - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
    9 Mjob - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
    10 Fjob - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
    11 reason - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other')
    12 guardian - student's guardian (nominal: 'mother', 'father' or 'other')
    13 traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
    14 studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
    15 failures - number of past class failures (numeric: n if 1<=n<3, else 4)
    16 schoolsup - extra educational support (binary: yes or no)
    17 famsup - family educational support (binary: yes or no)
    18 paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)
    19 activities - extra-curricular activities (binary: yes or no)
    20 nursery - attended nursery school (binary: yes or no)
    21 higher - wants to take higher education (binary: yes or no)
    22 internet - Internet access at home (binary: yes or no)
    23 romantic - with a romantic relationship (binary: yes or no)
    24 famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
    25 freetime - free time after school (numeric: from 1 - very low to 5 - very high)
    26 goout - going out with friends (numeric: from 1 - very low to 5 - very high)
    27 Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
    28 Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
    29 health - current health status (numeric: from 1 - very bad to 5 - very good)
    30 absences - number of school absences (numeric: from 0 to 93)

    # these grades are related with the course subject, Math or Portuguese:
    31 G1 - first period grade (numeric: from 0 to 20)
    31 G2 - second period grade (numeric: from 0 to 20)
    32 G3 - final grade (numeric: from 0 to 20, output target)

## **Selecting the columns & understanding the data**

-   After taking a look at the data. I will verify the suggestion from the news article that extra study time and paid tutors can support students academic performance, hence I will be using 2 columns: **studytime**; and **paid**.

-   For exploration, I will also examine the relationship of parents to students' performance with the following column: **Fedu, Medu, Fjob, Mjob, famrel, Pstatus**

-   I will only focus on the **G3** grade (final grade) to simplify the analysis

    For the explanation for each column, please check the previous section.

```{r}
#select the columns
student.data %>% select(school, studytime, paid, Medu, Fedu, famrel, Pstatus, G3)
```

# Exploratory Data Analysis

## **Correlation between school vs students' scores**

```{r}
boxplot1 <- boxplot(G3 ~ school, data = student.data, xlab = "School",
        ylab = "Students' Score", main = "School vs Scores", 
        varwidth = TRUE,
        col = "green", border = "Brown")

# Perform AOV to detect if the mean value between school is significantly
# different
test_school=aov(student.data$G3~student.data$school)
summary(test_school)
```

It looks that the students from GP school scores better than MS school. However, after checking with ANOVA test, the p-value = 0.372 \> alpha = 0.05), meaning there is no significant difference between the students' scores mean between schools. As such to compare other variables, I will use the data from both school.

## Correlation between duration of study and extra paid class

```{r}
##calculate studytime and paid classess mean
##group means of studytime
means_study <- aggregate(G3 ~ studytime, student.data,mean)
means_study2 <- means_study %>% mutate_at(vars(G3), funs(round(., 1)))
##group means of paid
means_paid <- aggregate(G3 ~ paid, student.data,mean)
means_paid2 <- means_paid %>% mutate_at(vars(G3), funs(round(., 1)))

##Plot Duration fo study vs Students' Score
a <- student.data %>% ggplot(aes(x = studytime, y = G3, fill=studytime)) + geom_boxplot() + theme_classic() + labs (x = "Study duration", y = "Students' Score") + stat_summary(fun=mean, colour="darkred", geom="point", shape=18, size=3, show.legend=FALSE) +
  geom_text(data=means_study2, aes(label = G3, y = G3 + 1), size=4) +
  scale_fill_discrete(name="Study duration", labels = c("1 = <2 hrs", "2 = 2 to 5 hrs", "3 = 3 to 5 hrs", "4 = >10hrs"))

##Plot Extra paid classes vs Students' Score
b <- student.data %>% ggplot(aes(x = paid, y = G3, fill=paid)) + geom_boxplot() + theme_classic() + labs (x = "Extra paid classes", y = "Students' Score") + stat_summary(fun=mean, colour="darkred", geom="point", shape=18, size=3, show.legend=FALSE) +
  geom_text(data=means_paid2, aes(label = G3, y = G3 + 1), size=4) + scale_fill_discrete(name="Extra paid classes")

plota <- ggarrange(a,b, ncol = 1, nrow = 1)
annotate_figure(a, top = text_grob("Study duration vs Students' Score", color = "Navy", face = "bold", size = 14))
annotate_figure(b, top = text_grob("Extra paid classes vs Students' Score", color = "Navy", face = "bold", size = 14))
```

If we compare the mean students' score of each studytime, it looks that the longer study duration will have some effect in students' score. So does extra paid classes. However, are the mean differences significant?

To make things more clear, I will do a Significance Test using one-way ANOVA to find if both study time and extra paid classes has any significant effect to the students' score.

```{r}
#Perform Significance Test
test_study <- aov(student.data$G3~student.data$studytime)
summary(test_study)
test_study2 <- aov(student.data$G3~student.data$paid)
summary(test_study2)
```

With one-way ANOVA done above, we can see that for study time vs scores value, it has Pvalue = 0.161 \> alpha = 0.05. Meaning we can reject the null hypothesis which says that the means between groups (in this case study duration) is the same. **Study time has no significant effect with students' academic performance**. Secondly, for the extra paid classes, Pvalue = 0.0428 \< alpha = 0.05. Meaning **extra paid classes has significant effect to students' academic performance**.

It is a slight surprise to see that study time doesn't really affect students' score, however extra paid classes/tutor can likely help students' scores. This is probably because there are a lot of factors that can affect students' academic performance. Hence why we will explore other variables.

## **Correlation between Parents Education, Occupation, Cohabitation, and Parents-Student Relationship Quality to Students' Score**

```{r}
#calculate Fjob and Mjob group mean
##group means of mother's job
means_medu <- aggregate(G3 ~ Medu, student.data,mean)
means_medu2 <- means_medu %>% mutate_at(vars(G3), funs(round(., 1)))
##group means of father's job
means_fedu <- aggregate(G3 ~ Fedu, student.data,mean)
means_fedu2 <- means_fedu %>% mutate_at(vars(G3), funs(round(., 1)))

#Plot into boxplot
##Plot Mom's Education vs Students' Score
a <- student.data %>% ggplot(aes(x = Medu, y = G3, fill=Medu)) + 
  geom_boxplot() + 
  theme_classic() + 
  guides(fill="none") +
  labs (x = "Mother's Education", y = "Students' Score") +
  # stat_summary(fun.data = get_box_stats, geom = "text", hjust = 0.5, vjust = 0.9)
  stat_summary(fun=mean, colour="darkred", geom="point", shape=18, size=3, show.legend=FALSE) +
  geom_text(data=means_medu2, aes(label = G3, y = G3 + 1), size=4)

##Plot Father's Education vs Students' Score
b <- student.data %>% ggplot(aes(x = Fedu, y = G3, fill=Fedu)) + 
  geom_boxplot() + 
  theme_classic() + 
  labs (x = "Father's Education", y = "Students' Score") +
  scale_fill_discrete(name="Parents Education", labels = str_wrap(c("0 = None", "1 = Primary (4th grade)", "2 = 5th-9th grade", "3 = Secondary", "4 = Higher Education"),15)) + 
  stat_summary(fun=mean, colour="darkred", geom="point", shape=18, size=3, show.legend=FALSE) +
  geom_text(data=means_fedu2, aes(label = G3, y = G3 + 1), size=4)

plot1 <- ggarrange(a,b, ncol = 2, nrow = 1, common.legend = TRUE, legend="bottom")
annotate_figure(plot1, top = text_grob("Parents' education vs Students' score", color = "Navy", face = "bold", size = 14))

#calculate Fjob and Mjob group mean
##group means of mother's job
means_mjob <- aggregate(G3 ~ Mjob, student.data,mean)
means_mjob2 <- means_mjob %>% mutate_at(vars(G3), funs(round(., 1)))
##group means of father's job
means_fjob <- aggregate(G3 ~ Fjob, student.data,mean)
means_fjob2 <- means_fjob %>% mutate_at(vars(G3), funs(round(., 1)))

##Plot Mother's Occupation vs Students' Score
c <- student.data %>% ggplot(aes(x = Mjob, y = G3, fill=Mjob)) + 
  geom_boxplot() + 
  theme_classic() + 
  guides(fill="none") + 
  labs (x = "Mother's Occupation", y = "Students' Score") +
  stat_summary(fun=mean, colour="darkred", geom="point", shape=18, size=3, show.legend=FALSE) +
  geom_text(data=means_mjob2, aes(label = G3, y = G3 + 1), size=4)

##Plot Father's Occupation vs Students' Score
d <- student.data %>% ggplot(aes(x = Fjob, y = G3, fill=Fjob)) + 
  geom_boxplot() + 
  theme_classic() + 
  labs (x = "Father's Occupation", y = "Students' Score") +
  scale_fill_discrete(name="Mother's/Father's Occupation") +
  stat_summary(fun=mean, colour="darkred", geom="point", shape=18, size=3, show.legend=FALSE) +
  geom_text(data=means_fjob2, aes(label = G3, y = G3 + 1), size=4)

plot2 <- ggarrange(c,d, nrow = 1, common.legend = TRUE, legend="bottom")
annotate_figure(plot2, top = text_grob("Parents' occupation vs Students' score", color = "Navy", face = "bold", size = 14))

#calculate famrel and Pstatus group mean
##group means of mother's job
means_famrel <- aggregate(G3 ~ famrel, student.data,mean)
means_famrel2 <- means_famrel %>% mutate_at(vars(G3), funs(round(., 1)))
##group means of father's job
means_pstatus <- aggregate(G3 ~ Pstatus, student.data,mean)
means_pstatus2 <- means_pstatus %>% mutate_at(vars(G3), funs(round(., 1)))

##Plot Father's Education vs Students' Score
e <- student.data %>% ggplot(aes(x = famrel, y = G3, fill=famrel)) + 
  geom_boxplot() + 
  theme_classic() + 
  labs (x = "Family Relationship Quality \n (1 = Bad - 5 = Excellent)", y = "Students' Score") + 
  theme(axis.title.x = element_text(size = 10)) + guides(fill="none") +
  stat_summary(fun=mean, colour="darkred", geom="point", shape=18, size=3, show.legend=FALSE) +
  geom_text(data=means_famrel2, aes(label = G3, y = G3 + 1), size=4)

##Plot Father's Education vs Students' Score
f <- student.data %>% ggplot(aes(x = Pstatus, y = G3, fill=Pstatus)) + 
  geom_boxplot() + 
  theme_classic() + 
  labs (x = "Parents' Cohabitation", y = "Students' Score") + 
  guides(fill="none") + scale_x_discrete(labels=c("A" = "Away", "T" = "Together")) + 
  stat_summary(fun=mean, colour="darkred", geom="point", shape=18, size=3, show.legend=FALSE) +
  geom_text(data=means_pstatus2, aes(label = G3, y = G3 + 1), size=4)


plot3 <- ggarrange(e,f,widths = c(1,1))
annotate_figure(plot3, top = text_grob("Students' score by family relationship & parents cohabitation", color = "Navy", face = "bold", size = 14))
```

**Parents education to students' score**

We can see that the higher the parents education, the students score also tends to get higher. Except for students with parents that has no education. However for "0" education parents, these only has a sample count of 2-3 students score.

```{r}
fedu.length <- aggregate(G3 ~ Fedu, student.data, length)
fedu.length
medu.length <- aggregate(G3 ~ Medu, student.data, length)
medu.length
```

**Parents occupation to students' score**

Different education also shows varying mean students' score, however are these difference significant?

**Family relationship quality to students' score**

We can see that a better family relationship suggest better students' score, as for the data with family relationship quality "1" only has 8 sample if we check below:

```{r}
famrel.length <- aggregate(G3 ~ famrel, student.data, length)
famrel.length
```

Hence it is hard to see the significance between variable.

**Parents' cohabitation to students' score**

The plot suggest that on average, students score is better if the students' isn't living in the same house with the parents.

### Significance Testing

To prove if one categorical variables (parents' education, occupation, family relationship quality, and parents' cohabitation) has any effect to students' scores, we will do one-way ANOVA to check the P-value for each comparison.

```{r}
# Perform one-way ANOVA

#Parents' education
##Mother's education
test.edu1=aov(student.data$G3~student.data$Medu)
summary(test.edu1)
##Father's education
test.edu2=aov(student.data$G3~student.data$Fedu)
summary(test.edu2)

#Parents' occupation
##Mother's occupation
test.ocu1=aov(student.data$G3~student.data$Mjob)
summary(test.ocu1)
##Father's education
test.ocu2=aov(student.data$G3~student.data$Fjob)
summary(test.ocu2)

##Family relationship
test.famrel=aov(student.data$G3~student.data$famrel)
summary(test.famrel)
##Parents' cohabitation
test.pstat=aov(student.data$G3~student.data$Pstatus)
summary(test.pstat)
```

The result of the Pvalue are:

> **Medu Pvalue = 9.24e-05 \< alpha = 0.05**
>
> **Fedu Pvalue = 0.0222 \< alpha = 0.05**
>
> Mjob Pvalue = 0.00519 \> alpha = 0.05 (reject null hypothesis)
>
> Fjob Pvalue = 0.268 \> alpha = 0.05 (reject null hypothesis)
>
> famrel Pvalue = 0.81 \> alpha = 0.05 (reject null hypothesis)
>
> Pstatus Pvalue = 0.25 \> alpha = 0.05 (reject null hypothesis)

Pvalue over alpha means we can reject the null hypothesis, meaning the variables doesn't have any significance towards students score.

Pvalue under alpha means we cannot reject the null hypothesis, meaning the variables has effect/correlates with students' score.

From this exploration that we have finished, we got 3 variables that correlates/has significant impact to students' scores, these are: **paid (extra paid classes), medu (Mother's education), Fedu (father's education)**

# Conclusion

1.  Study time has no significant effect on students' score, however this can probably be attributed to the possibility that a students' score is affected by many different variables. Extra paid classes seems to indeed correlates with students' score. It can be suggested that it is worth investing extra paid classes to help students' score.
2.  Both mother's and father's education correlates with students' score, it seems that better educated parents' can help students' score. However, other factor such as parents' occupation and parents cohabitation doesn't seem to correlates with students' score. Again, this can be attributed towards the possibility that a students' score can be affected by many factors.
